---
title: "cp-lf2023"
author: "DDB Perrakis"
date: "`r Sys.Date()`"
output: word_document
bibliography: references.bib
---

##[Get data]

```{r get.data, include=FALSE}
.libPaths("C:/Dan/RPackages")

knitr::opts_chunk$set(echo = FALSE, root.dir = "c:/Dan/_Remote_projects/ccp_lf/files")

rm(list=ls())
.libPaths("C:/Dan/RPackages")

require(stringr)
require(dplyr)
require(ggplot2)
require(lubridate)

load(file='./mcF.rda')
load(file='./wbmc5.rda')
load(file='./ISI.rda')

#get fd4 from cfi_process
#fd4 <- read.csv('./fire_data_current.csv') %>%
#  mutate(CFI2=factor(CFI, labels=c('Surface', 'Crown')))
fd4 <- read.csv('c:/Dan/_Remote_projects/ccp-lf/files/fire_data_oct2023.csv') %>%  mutate(CFI2=factor(CFI1, labels=c('Surface', 'Crown')))

#Deciduous data
d1 <- read.csv('c:/Dan/_Remote_projects/ccp-lf/files/fire_data_d1_oct2023.csv')

```

##[Create DFs]

```{r create data.frames, include=FALSE}
#Combine conifer data with D1
fd.all <- bind_rows(fd4 %>% select(
  num:RH, Grass.Cure) %>%
    mutate(Date=ymd(Date)), 
  d1 %>% select(num:RH, Grass.Cure) %>%
    mutate(Date=ymd(Date)))  

setwd('c:/Dan/_Remote_projects/ccp-lf')

sharp <- filter(fd4, str_detect(ExpProject, "Sharpsand IM")) %>% pull(num)  #original Sharpsand Immature stands only
sharp.th <- filter(fd4, str_detect(ExpProject, "Sharpsand TH")) %>% pull(num)
sharp.sm <- filter(fd4, str_detect(ExpProject, "Sharpsand SM")) %>% pull(num)

#get sharpsand (IM and TH) stand data
sh.fires0 <- filter(fd4, num %in% c(sharp, sharp.th, sharp.sm)) %>% 
  select(fire, Date, num, ws, MC.SA, MC.SA_dens, MC.SA_season, FMC, 
         FFMC, DMC, ISI, FSG1, SFC, ROS, 
         DC, FWI, CFI1, CFI2, Site) %>%
  mutate(Exp=case_when(
           num %in% sharp ~'Imm',
           num %in% sharp.th ~'Thin',
           num %in% sharp.sm ~'Semi-M' ),
         Plot=str_extract(fire, '\\d+') %>% as.integer) #extract num. from string

sh.fires1 <- mutate(rowwise(sh.fires0), 
                    MC.SA2=wbmc5(ffmc=FFMC, dmc=DMC, stand=4,                                           density=2, season=MC.SA_season)) #MC.SA_dens=2

#alternate isi, using mcSA and wind speed:
isi.mcsa <- function(mc, ws) {
  #m = 147.2*(101-ffmc)/(59.5+ffmc)  #original mc from FFMC (mcF) formula
  f_w = exp(0.05039*ws)
  f_f = (91.9*exp(-0.1386*mc))*(1+(mc^5.31)/(4.93*10^7))
  return(0.208*f_w*f_f)
}


#main dataframe for threshold analysis - Sharpsand fires
sh.fires <- mutate(rowwise(sh.fires1), 
                isi.m=isi.mcsa(mc=MC.SA, ws=ws),
                isi.m2=isi.mcsa(mc=MC.SA2, ws=ws))

```

##[SROS models]
```{r sros, include=FALSE}

#original ccp (2019) sf formula
sros.st <- function(isi) {
  0.6308 + 0.056*isi + 0.0086*isi^2
}


#surface fires
s.fires <- mutate(rowwise(fd.all), 
                  isi.m=isi.mcsa(mc=MC.SA, ws=ws)) %>% #get ISI using MC.SA
  filter(CFI1==0) %>% 
  mutate(FT=as.factor(case_when(
           str_detect(ExpProject, 'D1') ~ 'Decid',
           str_detect(ExpProject, 'Dewdrop') ~ 'PPDF',
           TRUE ~ 'Con'))
         ) %>%
  ungroup() %>% as.data.frame()
 
#Add FBP database C4/C5 missing VW fires
s.miss0 <- data.frame(num=122:125, fire=c('PNFI #1 13 yr ONT', 
                                          'PNFI #2 12 yr ONT', 
                                         'PNFI RW #1-41-2', 'Carp Lk PNFI WF QC'),
                     CFI1=0, CFI2='Surface', Fire.type='S', 
                     ROS=c(7.6, 4.3, 3, 7.5),
                     SFC=c(1.1, 0.59, 0.6, 0.9), CFC=c(0, 1, 0, 0),
      #Note estimated SFC for 3 of the 4
                     Date=c('1977-05-19', '1976-08-04', '1976-05-10',
                            '1963-07-03') %>% ymd(),
                     ws=c(19, 10, 18, 25), FFMC=c(91.7, 89.9, 90.4, 90.7), 
                     ISI=c(14.2, 7, 11.2, 16.7), DMC=c(57, 26, 20, 45), 
      DC=c(107, 276, 86, 81), 
                     BUI=c(57, 42, 25, 45), 
                     MC.SA_season=c(1, 2, 1, 2), MC.SA_dens=2, 
      MC.SA_stand=4, FT='Con')

#with corrected ISI values for the first two C4 PNFI fires
s.miss <- mutate(rowwise(s.miss0), 
                 MC.SA=wbmc5(FFMC, DMC, MC.SA_stand, MC.SA_dens, MC.SA_season),
                 isi.m=isi.mcsa(mc=MC.SA, ws=ws)) %>%
  as.data.frame() 
  

s.fires2 <- bind_rows(s.fires, s.miss)

#Original unaltered conifer SF only
s.fires.con <- filter(s.fires2, FT!='Decid') 

#conifer-only isi.m^2 ROS model
con.lm <- lm(data=s.fires.con, formula=ROS~I(isi.m^2)-1)

#prediction function for simple lm using isi.m
con.mod <- function(isim) {
  predict(con.lm, newdata=list(isi.m=isim))
}

#Noted - fires with some torching seem to be anomalous
#Exceptions: fires with torching, CFC >1
#except <- filter(s.fires, CFC >0.1)

#show exceptions, other issues
# ggplot(s.fires, aes(x=ISI, y=ROS)) + 
#  geom_point() + geom_point(data=except, shape=21, size=4)

#filter out fires with CFC; somewhat anomalous
#first convert all NA CFC values to 0
s.fires3 <- s.fires2 %>% 
  mutate(CFC=ifelse(is.na(CFC), 0, CFC)) %>%
            filter(CFC <= 0.1)

#convert ROS to presumed cured grass value (half-grass influence)
#Should work for cf=100 and cf=other (e.g. 90%)
c7.fullcuring <- 90  #
fc.cf <- 0.176 +0.02 * (c7.fullcuring-58.8)

s.fires.mod <- s.fires3 %>%
  mutate(CF=ifelse(Grass.Cure > 58.8, 0.176+0.02 * (Grass.Cure-58.8),
                   0.005 * exp(0.061 * Grass.Cure)),
         ROS=ifelse(is.na(Grass.Cure), ROS,
                    ROS+ROS*fc.cf*(1/CF)/2)) #actual ROS plus half of 'full curing' ROS if it was all grass


#surface fires, with false detrended C7; colour by FT
#with isi.m squared model and linear (con.mod)
ggplot(s.fires.mod %>% filter(!is.na(SFC)), 
       aes(x=isi.m, y=ROS)) +  
  geom_point(aes(colour=FT, size=SFC)) +    
  stat_smooth(method=lm, formula=y~I(x^2)-1, 
              se=FALSE, colour='black') +    
#  geom_point(data=s.fires.mod %>% filter(FT=='PDF'), shape=21, size=4) +
  stat_smooth(data=s.fires.con, method=lm, 
              formula=y~x-1, se=FALSE, colour='gray')
#Add dewdrop fires again untransformed

#Shown using ISI:
ggplot(s.fires.mod %>% filter(!is.na(SFC)), 
       aes(x=ISI, y=ROS)) +  
  geom_point(aes(colour=FT, size=SFC)) +    
  stat_smooth(method=lm, formula=y~I(x^2)-1, 
              se=FALSE, colour='black') +    
#  geom_point(data=s.fires.mod %>% filter(FT=='PDF'), shape=21, size=4) +
  stat_smooth(data=s.fires.con, method=lm, 
              formula=y~x-1, se=FALSE, colour='gray') 


#So far these models involve - 
# - Cutting out data with CFC > 1
# - 'Detrending' PIPO fires to simulate 100% grass curing, half
# - including substantial percentage of D1 fires

#linear model using ISI
sros.ISI2.mod <- lm(ROS ~ I(ISI^2)-1, data=s.fires.mod)
#r^2=0.7217 with s.fires3
#r^2=0.690 with s.fires.mod

#lm using isi.m
sros.isim2.mod <- lm(ROS ~ I(isi.m^2)-1, data=s.fires.mod)
#r^2=0.708 with s.fires2
#r^2=0.766 with s.fires.mod

#lm using isi.m AND SFC
sros.isim2SFC.mod <- lm(ROS ~ I(isi.m^2)+SFC-1, data=s.fires.mod)
#r^2=0.749 with s.fires2
#r^2=0.807 with s.fires.mod

#similar model with ISI (sros.ISI2SFC.mod) is also good, but not as good
sros.ISI2SFC.mod <- lm(ROS ~ I(ISI^2)+SFC-1, data=s.fires.mod)
#r^2=0.7407


sros.lin <- function(ws) {
  predict(sros.lin.mod, newdata=list(ws=ws))
}

#redo SF plot with new bivariate model



#first, fake 

#fake data for lines
f.sfires0 <- data.frame(fisim=seq(0:40), fsfc1=1, fsfc2=2)
f.sfires <-mutate(rowwise(f.sfires0),
                       fros1=predict(sros.isim2SFC.mod, 
                                     newdata=list(isi.m=fisim, SFC=fsfc1)),
                       fros2=predict(sros.isim2SFC.mod, 
                                     newdata=list(isi.m=fisim, SFC=fsfc2)))

#sigmoidal curve like FBP; try c6s to start
#then vary b and c params
c6s <- function(isi) {
  ros=30 * (1-exp(-0.08*isi))^3
  return(ros)
}

s.nl <- nls(data=s.fires.mod, formula=
               ROS ~ 25 * (1-exp(-b*isi.m))^c, start=list(b=0.08, c=3))

#just need to make a function
s.nls <- function(isi.m) {
  b=0.035177
  c=1.987496
  ros=25 * (1-exp(-b*isi.m))^c
  return(ros)
}

#plotted data and best model:
ggplot(s.fires.mod %>% filter(!is.na(SFC)), 
       aes(x=isi.m, y=ROS)) +  
  geom_point(aes(colour=FT, size=SFC)) +    
  #geom_point(data=s.fires.mod %>% filter(FT=='PDF'), shape=21, size=4) +
  geom_line(data=f.sfires, aes(x=fisim, y=fros1), colour='gray') +
  stat_function(fun=s.nls)
  


  

```

##[VW.lf mods]

```{r vw.lf mods, include=FALSE}
#Vw crown fire model-critical SFC
vw.sfc <- function(fmc=100, z, ros) {
  h = 460+25.9 * fmc
  I_0 = (0.01*h*z)^1.5
  #return(I_0)
  #Byram59
  sfc_0 = I_0/(300*ros)
  return(sfc_0)
  #critical sfc for crowning
  #use ros from surface ros model (VW or Perrakis 2020)
}

#VW model - critical intensity
vwi <- function(fmc=100, z) {
  h = 460+25.9 * fmc
  I_0 = (0.01*h*z)^1.5
  return(I_0)
}

#ladder fuel scaling
#Corrected!
lf.sfc <- function(zg, zl, fcl) {
  #zg is overstory LCBH
  #zl is ladder fuel height, or understory fuel centroid
  fcg=(zg/(zg-zl))^1.5 * fcl
  return(fcg)
}

#write.csv(lf.sfc, file='c:/Dan/_Fire_tools/r_scripts_functions/lf-sfc.rda')

sros.vw <- function(isi) {
 20*(1-exp(-0.2*isi))^5 
}

#Assume Sharpsand FSG ~ 4.5 m
#Snag height ~ 3.2 m
#with crown ratio of 0.5, dead CBH=1.6m, centroid = 3.2-(3.2 * 0.5 * 0.5)= 
zl.test = 4.5-3.2+(3.2 * 0.5 * 0.5)  #FSG_L



```


## Analysis

### #Sharpsand histograms with 3 treatments - hide

```{r Sharpsand Histograms}
scalex <- seq(2, 22, by=2)

#graph with histograms: ISI
sh.lad.histo <- ggplot(sh.fires, aes(x=ISI)) + 
  geom_histogram(binwidth=1, aes(fill=CFI2)) + 
  facet_grid(Exp~.) +
  scale_x_continuous(breaks=scalex) +
  theme_classic() +
  scale_fill_manual(values=c("#00AFBB", "#FC4E07")) +
  labs(title='Crown fire thresholds, Sharpsand Ck:ISI')

sh.lad.histo

#graph with histo: isi.m (mc.sa-based ISI, using season, density, DMC)
sh.lad.histo2 <- ggplot(sh.fires, aes(x=isi.m)) + 
  geom_histogram(binwidth=1, aes(fill=CFI2)) + 
  facet_grid(Exp~.) +
  scale_x_continuous(breaks=scalex) +
  theme_classic() +
  scale_fill_manual(values=c("#00AFBB", "#FC4E07")) +
  labs(title='Crown fire thresholds, Sharpsand Ck: ISI.mcsa')

#sh.lad.histo2
```

Crown fire thresholds for the 3 Sharpsand Creek stand structures are
shown in Figure 2. Both the site factor and ISI were significant
(alpha=0.05) in predicting crown fire occurrence.

Tried pooling sites together; probably not much there.

### #Pooling SM and TH, new histogram - hide

```{r Pooling thinned and semi-m trts}
fires3 <- mutate(fires2, 
                 Exp.pool = case_when(
                   Exp=='Imm' ~ 'Imm',
                   TRUE ~'Th and SM'
                 ))

#graph with pooled treatments, ISI
sh.lad.histo3 <- ggplot(fires3, aes(x=ISI)) + 
  geom_histogram(binwidth=1, aes(fill=CFI2)) + 
  facet_grid(Exp.pool~.) +
  scale_x_continuous(breaks=scalex) +
  theme_classic() +
  scale_fill_manual(values=c("#00AFBB", "#FC4E07")) +
  labs(title='Crown fire thresholds, Sharpsand Ck: ISI, pooled')

sh.lad.histo3
```



##Solving for M, the ladder fuel multiplier

We will work with the Immature and Thinned crown fire thresholds to solve for M, using a range of techniques, including expert opinion, comparing critical intensity for crowning between both stands, and using fitted ROS models.     


###Estimation 2: Using a surface ROS model with Byram's intensity
For the second estimation method, we will test M using Byram's I_0 thresholds applied to the portion of the experimental fire dataset representing the Sharpsand Creek fires. This site, with many fires in stands that differ primarily by age and treatment, provides the ideal test case for fitting a ladder fuel multiplier. 

Work with thresholds:

Immature ISI: `r isi.thresh.imm()` #6.61
Immature, ISI.mcsa: #4.70`

Thinned, ISI: `r isi.thresh.th()` #10.31
Thinned, ISI.mcsa: #9.80

Semi-mature, ISI: `r isi.thresh.semiM()` #8.62
SM, ISI.mcsa: #8.95



###Estimation 3: fitting M to empirical CFO model
We will refit the crown fire occurrence database, previously described in Perrakis et al. [@perrakis2023]. This time, in addition to the b_n coefficients, we will fit M, the ladder fuels multiplier, applied to FC_G values as a random variable. This includes testing the upper canopy crowning at the two sites with two-layered canopies (Kenshoe Lake and ICFME; [@perrakis2023]). 


[Delete? At mean WS_10 of 12.3 km /h, the IM threshold works out to FFMC of
about 88.7 and TH threshold is about FFMC 91.8. Using the MC.SA-based models and the same wind speed, the IM and TH thresholds work out to MC.SA values of 14.8 and 9.8, respectively.  ]

Third method: Byram's I_0 with empirically-derived crowning thresholds. 
Using Byram's [-@Byram1959] equation, we will use I_0 along with the two empirically-derived threshold ISI values for CFI to calibrate the SFC multiplier for ladder fuel consumption:

$$I_{0}=h \cdot SFC_0 \cdot ROS$$
We will use the FBP System conventions for units and constants: I~0~ in kW m\^{-1}, SFC in kg m\^{-2}, ROS in m min\^{-1}, and h=18,000 kJ kg\^{-1}; to convert ROS to m s\^{-1}, we multiply by the constant 1 min/60 s  [@FCFDG 1992]. We assume that the LCBH of both the Immature and Thinned stands are, for working purposes, both identical at about 4.4 m, since the experimental stands were located at the same site and differed only (to our knowledge) in the thinning treatment conducted ~15-20 years prior to burning. This is in line with a recent crown fire analysis estimating LCBH for both stands at 4.4-4.5 m [@perrakis2023] as well as previous studies using these data that characterized LCBH as identical (4 m) in both stand types [@stocks1987; @vanwagner1993; @FCFDG1992; @cruz2004]. Therefore, the critical surface fire intensity for engaging the 4.4 m LCBH canopy fuels, once the ladder fuels are accounted for, should be comparable. We will also assume that a similar sROS model could be used in both stands, though we will discuss and evaluate this assumption later. 

If $h$ is the same in both stands, then 
$$ ROS_{0.IM} \cdot SFC_{0.IM}=ROS_{0.TH} \cdot SFC_{0.TH}$$

In the IM stand, with mean SFC of 1.41 kg/m\^2, the crowning threshold was ISI `r isi.thresh.imm()` #6.61 or ISI.m #4.70. With a dead stem mean height of 3.49 m, the ladder fuel centroid $C_L=$ 1.75 m, and $z-z_L= 2.62 m$. Ladder fuel consumption, based on the loading of \< 1 cm dead roundwood [@stocks1987], was estimated at 0.159 kg/m2. Using equation 6, this scales to $FC_G= 0.3392 kg m^{-2}$. We now define the final SFC value including ladder fuels scaled to the ground as the effective SFC, where $SFC_e=SFC + FC_G \cdot M$ (actual SFC plus scaled-to-ground ladder fuel consumption times a multiplier). Thus, $SFC_e=1.41 + 0.3392 M$. 

In the TH stand, mean SFC was 1.12 and we assume the remaining dead ladder fuels represented a baseline, negligible effect. 
 
[We will use a simple fitted ROS model based on Canadian experimental fire data, previously presented during the creation of the 'Conifer Pyrometrics' system. This model uses $ISI$ and $ISI^2$ as predictors:
$$ROS_S=0.6308 + 0.056 \cdot ISI + 0.0086 \cdot ISI^2$$ ] Don't use!

[Use isi.mcsa model instead]

Substituting $SFC_e$ for $SFC_{0.IM}$ and solving for $M$ gives the following  algebraic solution:
$$SFC_e=\frac {ROS_{0.TH} \cdot SFC_{0.TH}} {ROS_{0.IM}}$$
and
M=[[ROS_0th . SFC_0TH / ROS_0IM] -1.41]/0.3392

[works with ROS_0 values below....]
$$M=\frac {\frac {ROS_{0.TH} \cdot SFC_{0.TH}} {ROS_{0.IM}}-1.41} {0.3392}$$


## Get CFI data for empirical analysis



[Not sure if this part below is to be included or not]

## Discussion

As Andrews [-@andrews2018] noted, Rothermel surface model only
applicable to surface fuels, within \~6ft of the ground

Thinning increases sub-canopy turbulence [@russell2018]. See also
Banerjee et al 2020

McAlpine and Xanthopoulos [@mcalpine1989] also noted that C-6s appeared
to overpredict sROS in needle fuelbed experiments. Although they
explained it as a difficulty in estimating

Method could also be used to estimate influence of dead branch fuel
below LCBH, bark flakes, arboreal lichens, or similar material.

```{r get stand data calculate FC_L}
sh.full <- read.csv('c:/Dan/_Remote_projects/ccp_git/ccp-cfi/sharp_full.csv') %>%
  mutate(FSG=LCBH) %>%
  select(-c(X, BA, BA.d, LCBH.L)) %>%
  left_join(fires %>% filter(Exp %in% c('Thin', 'Imm')), by='Plot') %>%
  select(-c(fire, Date, num, DC, FWI, CFI2, Exp)) 

lf.cfl <- c(0.142, 0.073, 0.313, 0.183, 0.133, 0.146, 0.131, 0.131,
              0.176, 0.182, 0.151, 0.218, 0.084) #p11 twice

#crown fuel consumption estimates by fire type
cfc.p <- c(0.39, 0.49, 0.22, 0.46)
cfc.c <- c(0.89, 1.16, 0.99, 1.27, 1.06, 1.4, 1.04, 1.03, 1.11)
cfc.s <- c(0, 0, 0.08, 0, 0)


#* 0.815# using woody consumption? use full CFL here I think; 
#ICFME surface woody consumption was 0.6 (Stocks et al 2004)
#ALso ICFME: Stocks et al (2004): 70-86% of overstory 0-1 cm roundwood were consumed, with moisture contents from 55.5-90.4 % 
#down woody fuel moisture contents were 6.3-14.5 %

#Although P7, 18 (ctrl) and P8, 9, 16 (th) are surface fires, assume dead snags burn
dead.cfc.mean <- mean(lf.cfl)

s.ha.ctrl <- filter(sh.full, TRT=='ctrl') %>% pull(s.ha.d)
dead.cfl.snag <- mean(lf.cfl*10000/s.ha.ctrl)  #mean snag weight cfl (<1 cm) per stem, from ctrl stands; in kg/stem

sh.ctrl <- filter(sh.full, TRT=='ctrl') %>% 
  mutate(cfc.d=lf.cfl,
         cfc.d.stem=cfc.d * 10000/s.ha.d,   #kg/m2*10000 m2/ha/s/ha = kg/stem
         z.l=FSG - HT.D + (HT.D * 0.5 * 0.5),  #assume CR of 0.5;
         sfc.add=cfc.d * (FSG/z.l)^1.5,  #SFC.2=(z.2/z.1)^1.5 * SFC.1
         sfc.add.stem=sfc.add/s.ha.d) 

sh.ctrl2 <- sh.ctrl %>% select(Plot, ISI, FSG, SFC, sfc.add, CFI=CFI1) %>% 
  mutate(sfc.both=SFC+sfc.add)

```

```{r}
#Get surface fires in mod or dense class(exclude GL)

fd.sf <- filter(fd4, CFI1==0 & num != 68 &
#               !num %in% c(sharp, sharp.sm, sharp.th) &
                MC.SA_dens >1) %>% as.data.frame()

#surface fire model, fitted to ISI for all pine mod/dense sf in database (excluding Sharpsand)
sf.mod <- lm(data=fd.sf, formula= ROS ~ I(ISI^2) -1)  #ISI^2 model, forced through 0; looks good
#adj. r^2=0.7207

sf.nl <- nls(data=fd.sf, formula = ROS ~ a + ISI^b, start=list(a=0, b=2))
#Nah - gives b= ~ 0.5, wrong curve shape

#97.5% CI for estimate - not really that good; need 95% for prediction, not param.
#sf98 <- confint.lm(sf.mod, level=0.95)[2] #level=0.8 gives 10% and 90% CI

sf.fun <- function(ISI) {
  sf.mod$coefficients[1]*ISI^2
}


#data for ribbon
ribbon.df <- data.frame(ISI=seq(min(fd.sf$ISI), max(fd.sf$ISI))) %>%
  mutate(ROS=sf.fun(ISI),
         ROS.d35=ROS*0.65, 
         ROS.u35=ROS*1.35)
                        
fd.sf.sharp <- filter(fd.sf, Site %in% c(10:12))  #Sharpsand only

#graph of surface fires, with VW ROS model and fitted ROS model, with ribbon
ggplot(data=fd.sf, aes(x=ISI, y=ROS))+ 
  scale_shape_manual(values=c(2, 5, 6, 0, 1)) +
  geom_point(aes(colour=ExpProject, shape=str_extract(ExpProject, '\\w+'))) +
  geom_point(data=fd.sf.sharp, shape=1, size=4) +
  #geom_point(aes(size=FSG, colour=str_extract(ExpProject, '\\w+'))) + #no not useful in the end
  #geom_smooth(method='lm', formula= 'y~poly(x,2)') +
  geom_ribbon(data=ribbon.df, aes(ymin=ROS.d35, ymax=ROS.u35),
              alpha=0.1) +
  stat_function(fun=ros.vw, colour='grey') +
  stat_function(fun=sf.fun, colour='black') +
  theme_classic() +
  labs(shape='Site') +
  scale_colour_discrete(guide='none') +
  annotate('segment', x=isi.thresh.imm(), xend=isi.thresh.imm(), 
           y=sf.fun(isi.thresh.imm())-1, 
           yend=sf.fun(isi.thresh.imm())+1, colour='red') +
  annotate('segment', x=isi.thresh.th(), xend=isi.thresh.th(), 
           y=sf.fun(isi.thresh.th())-1, 
           yend=sf.fun(isi.thresh.th())+1, colour='blue') +
  geom_text(x=10, y=ros.vw(10)+3, label='VW SF model')

  
```

```{r}
#Mean absolute error - SF model
fd.sf.mae <- mutate(fd.sf, 
                      ROSvw=ros.vw(ISI),
                      ROSsq=sf.fun(ISI),
                      abserr.vw=abs(ROSvw-ROS),
                      abserr.sq=abs(ROSsq-ROS)) %>%
  summarize(mae.vw=mean(abserr.vw),
            mae.sq=mean(abserr.sq))
#mae.vw=4.400
#mae.sq=0.757

#create table of predicted/observed I_0 using vw and new squared ISI model
sh.ctrl2row <- mutate(sh.ctrl2, 
                      Ivw = vw.sfc(fmc=111.5, z=FSG, ros=ros.vw(ISI)),
                      Isq = vw.sfc(fmc=111.5, z=FSG, ros=sf.fun(ISI)))

cfc.dstem <- summarize(sh.ctrl, mean(cfc.d.stem)) %>% pull()#cfc per dead stem, kg

sh.th <- filter(sh.full, TRT=='th') %>% 
  mutate(cfc.d.stem=cfc.dstem, #use cfc in ctrl plots to estimate cfc in th plots
         cfc.d=cfc.d.stem * s.ha.d / 10000, #kg * s/ha / 10000m^2/ha to give kg/m^2
         z.l=FSG - HT.D + (HT.D * 0.5 * 0.5),  #assume CR of 0.5; crown cent. FSG
         sfc.add=cfc.d * (FSG/z.l)^1.5,  #SFC.2=(z.2/z.1)^1.5 * SFC.1
         sfc.add.stem=sfc.add/s.ha.d) 

 sh.th2 <- sh.th %>% select(Plot, ISI, FSG, SFC, sfc.add, CFI) %>% 
  mutate(sfc.both=SFC+sfc.add)

sh.th2row <- mutate(sh.th2, 
                      Ivw = vw.sfc(fmc=111.5, z=FSG, ros=ros.vw(ISI)),
                      Isq = vw.sfc(fmc=111.5, z=FSG, ros=sf.fun(ISI)))

sh.both <- full_join(sh.ctrl2row, sh.th2row) %>%
  mutate(vw.cfi.sf = SFC > Ivw,
         vw.cfi.both = sfc.both > Ivw,
         sq.cfi.both = sfc.both > Isq,
         sfc.m=sfc.add * 3.25 + SFC,
         sq.cfi.m = sfc.m > Isq)
#Threshold ISI fires just enough to get over 50% prob. of CFI
#calc. at mean SFC
```

The problem with a semi-physical assessment of the Sharpsand Creek fires
is that the fundamental relationship appears shaky based on the terms
originally provided. The surface fire model proposed by Van Wagner
[-@vanwagner1993a; 'to define the spread rate of all possible surface
fires'] for immature jack pine fires (i.e. control stands at Sharpsand
Creek) is shown in Figure X to compare with empirical data. The figure
also shows the ISI-ROS relationships for 37 experimental fires from
closed pine stands at various sites across Canada (Supplemental Table
S1). A simple empirical model was fitted to these observations:

$$
ROS_S=0.02214 \cdot ISI^2
$$

[6], with adjusted $R^2$ of 0.706 and mean absolute error (MAE) of
`r  fd.sf.mae[2]`.

This is a much slower surface ROS model than that originally provided,
with scant evidence, by Van Wagner's Equation 5. Although the data shown
in Figure X represents fires from several different sites across Canada,
it strongly suggests that surface fire under a closed pine canopy is
generally much slower than previously suggested. Furthermore, it is now
understood that the surface fuels contributing to an upward heat flux
during flaming combustion in a typical boreal conifer stand are not
comprised of the entire surface fuel layer, but rather only the
well-aerated litter and fine woody debris. These two "errors" were
self-correcting in Van Wagner's original description, but can now be
amended. Instead, if we use the fitted model from Equation 6 to predict
surface ROS at a given ISI value, we are left with much higher required
$SFC_0$ values than observed.

Table X, below, shows the Immature Sharpsand fires, along with estimated
$SFC_L$ calculated from Equation 5.

```{r}
sh.both %>% round(3) %>% 
  mutate(pCFIvw=Ivw)
```

How much of the elevated SFC from ladder fuels was consumed?

The details of partial crowning in a stand are virtually impossible to
predict except probabilistically (e.g. de Groot et al. 2022). A surface
fire moving through a stand is continually affected by wind gusts,
tempered by canopy influence (Moon et al. 2016) and variability in
surface fuelbed properties such as bulk density and loading (ref.
surface fuel var. #Keane?). When a crown fire is involving \< 50% of the
canopy fuels, we might still consider the fire to be a passive (e.g.,
Van Wagner 1993) or intermittent (e.g. FCFDG 1992) crown fire, but
cannot know which elements of the ladder fuel layer will be consumed, or
will result in flame transmission to the live overstory. Such patchy
fire behaviour can be part of fire acceleration from point source
(McAlpine and Wakimoto 1991), but when it is reached under equilibrium
conditions, the upward heat flux from individual ladder fuel elements
(living or dead trees, etc.), clumps or 'jackpots' of surface fuel, etc.
is a probabilistic process. By including and scaling all of the ladder
fuel mass in the $FC_L$, the potential exists to overpredict the ladder
fuel influence, but this

# .
# .

# Headings
# Simple solutions for modelling surface fire and ladder fuels in Canadian conifer forests
Daniel D. B. Perrakis
Dan K. Thompson

##Abstract
Predicting the onset or occurrence of conifer crown fires is an uncertain process. Several operational modelling systems rely on the C.E. Van Wagner (Can. J. For Res. 1977, v7: 23-34) crown fire initiation model derived from simple physical theory and laboratory experiments. Van Wagner's model requires a clear vertical separation between surface and canopy fuel strata, and defines fine-textured biomass located between strata as ladder fuels (LF). While the LF layer contribution to crown fire initiation is not quantified by Van Wagner's model or in most prediction systems, it is suspected to be disproportionately important. In this study, we show how a simple rearranging of Van Wagner's model reframes crown fire initiation as a critical fuel threshold-driven process. A new equation is presented for comparing fuel consumption at different vertical positions in the mid-story, effectively treating LF consumption as though these fuels were lowered to the ground level and boosted in mass. We used fire observations and crown fire thresholds from a well-documented series of experimental burns in Canadian conifer forests to fit simple surface fire models and test the thoretical LF scaling equation. Incorporating LF into empirical crown fire models improved mode results; best results were obtained with a LF multiplier of [??2-3??; draft results], suggesting that LF consumption is about two to three times as influential per unit mass as typical conifer surface fuel consumption. Resulting models will be useful for both first estimates of surface fire spread rate as well as for incorporating LF into separate crown fire models or hazard reduction treatment design. 

##Introduction
Crown fires, CEVW model and ladder fuels; short lit. review (Stephens' papers, lab studies, etc.)
Conifer forests cover most of the circumboreal region of the world and
crown fires represent a significant fire-related hazard in these stands
[@kneeshaw2011, @stocks1991]. As noted by Kiil et al. [@kiil1976] and quoted by Cruz et al. [@cruz2005], surface fires rarely spread \> 6 m/min under closed conifer stands, as higher spread
rates are usually associated with some degree of crowning or torching.
Once flames reach the upper reaches of the canopy, they can be directly
influenced by open air winds in the lower atmosphere. The fully developed crowning process has long been known to be associated with a dramatic increase in fire spread rate, due to several processes including flame tilt, increased lee-side wind exposure [@taylor2004] and vigorous ember spotting [@cruz2016].

Stand structure is well-known to influence crown fire occurrence and tendency. The element in fuel complexes that enables crown fires in most tall
conifer stands are ladder fuels. This term refers to any easily-ignited
fuels in the mid-canopy space, meaning fine-textured and flammable dead
or live biomass above the surface fuel complex but below the live canopy
base. In Canada, J.G. Wright first noted that 'crown fires are not
likely to occur unless there is a large volume of fuel under the trees'
[@wright1932]. Van Wagner [@vanwagner1977], in describing his classic
crown fire initiation model based on empirical data and physical
convection theory, noted that such 'bridge fuels' consisted of
'combustible matter such as loose bark, dead lower branches, lichen,
small conifers, etc.' located above the surface fuel complex, remarking
that such fuels needed to be 'present in sufficient quantity to
intensify the surface fire appreciably as well as to extend the flame
height'. While the paper describing Van Wagner's model has been cited become a
classic (> 1500 citations as of 2023), methods for quantifying the effects
of ladder fuels are still lacking 45 years later.

Van Wagner's model was originally formulated to solve for critical
surface intensity ( $I_0$) as follows:

$$
I_0=(chz)^{1.5}
$$

[1],

where $h$ is heat of ignition, $z$ is live crown base height (LCBH), and
$c$ is a dimensional constant derived empirically. Most modelling
systems have combined $I_0$ with Byram's [-@byram1959] fireline intensity equation in order to solve for critical rate of spread. This combination has allowed for linkages with Rothermel's [-@rothermel1972] semi-physical surface rate of spread model [e.g. @scott2001; @andrews2014], or with the empirical ROS functions
from the FBP System fuel types [@forestrycanadafiredangergroup1992].
Surprisingly, a simple algebraic rearranging of the terms of Van Wagner's model allows for a useful and powerful estimator applicable to ladder fuels of varying
heights.


### Objective

The objective of this study was to describe a simple theoretical framework for
analyzing ladder fuel influence in crown fire initiation, based on a rearrangement of Van Wagner's (1977) equation. Secondary objectives include calibrations and evaluations of the theoretical equation, as well as refitting empirical crown fire models for use with measured or estimated ladder fuel loadings and structure. 


### Theoretical model: scaling ladder fuel consumption to the surface

Ladder fuels are defined by their fine texture and intermediate vertical position, by definition [@vanwagner1977; @kilgore1975]. They exist in conifer stands between
the surface fuels, at $z=0$, and canopy fuels, which for our purposes begin at the live canopy base height (LCBH). When connected to the surface fuels and
ignited by a proximate source (i.e. an advancing fireline), the heat
flux contribution [@melnik2022] is much closer to the canopy base, and therefore much more likely to contribute to canopy ignition compared to similar fuel
loading at the surface. The structure of the Van Wagner (1977)
CFI model permits a vertical rescaling that relies on an implied equivalency between SFC and the inverse of z (LCBH).

When we replace $I_0$ (Equation 1) with Byram's (1959) fireline
intensity, we get the following equation, a much-used equivalency (e.g. FCFDG 1992, Scott and Reinhardt 2001):

$$
H \cdot SFC \cdot ROS=(c \cdot h)^{1.5} z^{1.5}  
$$[2]

where $H$ is the heat of combustion, $SFC$ is surface fuel consumption,
$ROS$ is rate of forward spread, and the right side is as in Equation 1.

As discussed, this relationship has often been evaluated in terms
of identifying critical ROS for crown fire initiation (Van Wagner 1977,
Alexander 1988). However, if we assume that surface ROS is insensitive
to small variations in LCBH, it should also hold for comparing a
range of SFC values. Under surface fire conditions near the crown fire
initiation threshold, critical SFC could then be defined as follows:

$$
SFC_0=\frac {(ch)^{1.5} z^{1.5}} {h \cdot ROS} 
$$

[3]

Critical SFC for crown fire initiation can then be compared between LCBH
levels, e.g. $z_1$ and $z_2$.

$$
\frac {SFC_2}{SFC_1}=
\frac {  \left[  \frac {(ch)^{1.5} z_2^{1.5}} {h \cdot ROS} \right] } 
{  \left[  \frac {(ch)^{1.5} z_1^{1.5}} {h \cdot ROS} \right] }
$$

[4]

This also assumes no change in ROS with varying LCBH. Holding all
other terms constant, we can simplify this to yield a basic relationship
between SFC and LCBH (z) at different heights:

$$
SFC_2=(\frac {z_2} {z_1})^{1.5} \cdot SFC_1
$$

[5].

A practical example helps illustrate the logic of comparing $SFC_0$ at two different z values (equation 3). Using Van Wagner's own empirical heat
of ignition function [@vanwagner1968], at 90% foliar moisture content
(FMC) and LCBH=2 m, Equation 1 suggests $I_0$ of about 417 kW/m, the
intensity of a low to moderate intensity surface fire. At some moderate
surface ROS value, e.g. 3 m/min, Equation 3 predicts $SFC_0$ of 0.463
$kg \cdot m^{-2}$. Using Equation 5, increasing LCBH to 5 m without
altering other inputs would raise $SFC_0$ to 1.83 $kg\cdot m^{-2}$ ;
thus, a 2.5 times increase in LCBH would result in a fourfold increase
in fuel consumption needed for crowning as a consequence of the much higher surface fire intensity required. 

In the case of the consumption of an elevated ladder fuel quantity
$FC_L$ positioned at height above ground $z_L$, such as the vertical centroid of
a cohort of standing dead trees, finding the equivalent $SFC$ value at
ground level ($FC_G$) is then a matter of inflating its influence as we
scale it vertically. In this case, z is the normal (ground-level) LCBH and $z-z_L$ can also be considered the fuel strata gap between ladder fuel elements and the LCBH:  

$$
FC_G=(\frac {z} {z-z_L})^{1.5} \cdot FC_L ~~, ~~~~~~ z_L > z
$$

[6]. 
Aside from Van Wagner's equation, this is simply a slightly different representation of Thomas' [-@thomas1963] equation for flame size under calm conditions. Using Equation 6, it becomes possible to estimate the contribution of
parts of the fuel complex based on their vertical position in the crown
as if they were part of the surface fuel complex.

We note immediately two fire-related qualifiers. First, this theoretical
relationship does not account for the obvious structural differences
between the surface and crown fuel complexes. In particular, surface
fuelbeds are much more compact (higher bulk density and packing ratio)
than crown fuel complexes, including ladder fuels, with a much more aeration-limited combustion
environment [@keane2015; @schwilk2015]. The general FBP System approach
to SFC, where all consumed surface and ground fuels are assumed to
contribute to surface fire intensity, may also overemphasize dense fuels
that burn in post-frontal combustion [#could use some more references
and discussion on this point]. In contrast, fine crown and ladder fuel strata
are much more aerated (more porous) but also vertically and
horizontally discontinuous (refs and explanation#). This difference
could be addressed theoretically or, as in the present investigation,
empirically.

Second, while equation 6 specifies that the ladder fuel height cannot
exceed the LCBH, it is apparent that as the difference between them
($z-z_L$) diminishes, $FC_G$ can grow to extreme levels. Thus, 0.1 $kg \cdot m^{-2}$ of fuel consumption 1 m below a 5 m tall canopy base scales to equate to ground-level SFC of 1.12 $kg \cdot m^{-2}$ , but the same fuel consumption 0.2 m below the same canopy scales to $FC_G=12.5~ kg \cdot m^{-2}$. The logic of this holds in a theoretical sense -- a fire is much more likely to bathe the lower canopy in flame if the burning fuel is very close below, and a 0.2 m separation is essentially showing negligible separation at the fuel particle scale; however, this suggests that caution is needed to utilize eq. 6 in a practical sense to avoid inflating the ladder fuel effect to illogical levels.

The remainder of the study tests and discusses this theoretical relationship using previously documented Canadian experimental fire data. 

##Methods

Three separate validation methods were used to evaluate the theoretical ladder fuel scaling relationship: 1) an expert opinion estimate; 2) a logistic model comparing treatment types; and 3) and empirical model  All of these relate to the stand level

### Study area: Sharpsand Creek, Ontario

Stand and fuel measurements were initiated in the early 1970's at this site north of Thessalon, Ontario, in advance of experimental burns conducted over a 15-year period. This resulted in 27 separate and independent experimental fire observations in three
distinct structure classes. The original unaltered stand (immature: IM) consisted
of densely stocked jack pine natural regenerated on gravelly soil
[@walker1975a] following the spring 1948 Mississagi fire [@stocks1973].
The dense stands in the earliest burns featured over 9000 live stems/ha
and over 10,000 dead stems/ha as the stand underwent rapid
stem-exclusion. Results from the earlier fires (1975-1981) were
published in Stocks [-@stocks1987a] and included in the empirical
derivation of the C-4 fuel type model [@fcfdg1992]; these were also
analyzed separately by Van Wagner [@vanwagner1993] and discussed in
several other fire behaviour studies [@call1997; @cruz2003; @albini1986].

Results from the later fires (1988-1991) were not included in the FBP
System analyses and were fully documented only recently [@mcrae2017a].
The stand characteristics associated with the later experimental fires
highlighted significant structural and successional changes underway as
stands evolved: the dense immature forest was rapidly self-thinning,
from 9,276 live stems/ha at the time of the original experiments
(measured in 1973) to less than half the live density (4,375 stems/ha)
as well as 53% more basal area in 1984. These were termed 'semi-mature'
(SM), and estimated to also have average LCBH of 5.3 m, about 1 m higher
than the IM stands. During the course of the earlier burn experiments,
it was discovered that 6 of the plots had been hand-thinned (TH) in
1960, removing most of the live stems and resulting in much less dense
stands compared with the IM plots [@stocks1987a; B. J. Stocks, personal
communication, Jan. 2020]. While these thinned stands were the same age
and site characteristics as the denser IM (C-4 fuel type) stands, the burns in the
thinned plots had been deemed more representative of mature pine stands
and previously included in the FBP System C-3 fuel type dataset
[@degroot2022].

Burning years ranged from in 1974-1976, 1981, 1988, 1990-1991; thus, at
post-fire ages 26-43. As discussed in a previous analysis [@Perrakis2023], earlier studies handled these three treatments in different ways, but mainly by
estimating the ladder fuel influence as a modification of the LCBH in a
subjective sense, using an expert judgment adjustment and defining the fuel strata gap (FSG) as the net distance between surface and fuel strata, including the effects of standing dead ladder fuels [@cruz2004, @cruz2006, @perrakis2023].

There is a question of which plots experienced crown fire. In the
original Sharpsand experiments (1974-1981; Stocks 1987; Cruz 1999), fire
type was identified as 'surface' or 'crown'. In the latter experiments,
however, fire type was identified as 'surface', 'some torching',
'torching', and 'crown' [@mcrae2017]. Descriptions in the text make it
clear 'torching' fires experienced passive crowning, while those
described as 'some torching' appeared to be near the crowning threshold
(i.e., \~5-15% crown fraction burned). Following discussions with
experimental leaders (BJ Stocks, pers. comm.), the two fires described
as 'some torching' were split, with fire #16 (ISI=8.1, ROS=3.5 m/min)
was considered a surface fire, while fire #18 (ISI=9.4, ROS=5.1 m/min)
was considered a passive crown fire. As noted in a previous study, as surface fires begin to involve more canopy fuels and transition to passive crown fire, ROS is observed to increase beyond the baseline for true surface fires (i.e., no crown fuel involvement).

Thinned: ISI threshold about 10.2 (above 11.4, all CF; below 10.4 all
SF) =ws17, FFMC90.1; mc.ffmc=10.73

Overall mean SFC 1.3

The most challenging value for the analysis is the fuel consumption of
ladder fuels, as this is seldom measured (though sometimes estimated
indirectly) in experimental studies. Stocks [-@stocks1987] presented
estimated dead roundwood fuel consumption for the original Sharpsand
Creek exprerimental fires, and these values are used in the present analysis.

For the calculations, ladder fuel influence was incorporated into
certain models using equation 6 to scale the fine dead ladder fuels
(fine snags and dead crown fuel \< 1 cm; Stocks 1987) into an equivalent SFC
contribution value, SFC_L. The vertical position of fine standing ladder
fuels was assumed to be the crown centroid of the mean dead snag height
[LCBH+(height-LCBH)/2; @perrakis2023] in an experimental plot (Figure
X#snag diagram). SFC_L was then added to the actual (measured) SFC for
modelling. Additional tests were completed where the elevated fuel
contribution was increased using a multiplier, discussed further below.

###Estimation 1 - expert opinion
The first method involved expert opinion from previous studies. Cruz et al. [-@cruz2004] noted that the abundant ladder fuels characterizing the IM stands, in the form of standing dead trees (including dead branches, bark flakes, and fine stemwood) amounted to a reduced FSG of 2 m, compared to approximately 4 m for the TH stands. 

We used those values as a first estimate along with equation 6 to investigate a potential value for multiplier M. 

$$\begin{aligned}
M
&=(\frac {4} {2})^{1.5} \\
M&=2.828 \end{aligned}$$
[7]. Instead of those approximate values, we could also use values derived from the more precise estimates from original stand data [@walker1975]. As previously described [@perrakis2023], detailed stand survey analysis produced mean treatment-level LCBH values of 4.45 m (TH) and 4.36 m (IM). If we assume that the 'expert opinion' assessment of a 2 m reduction in FSG in the IM was correct, we get the following values for equation 6: 
$$\begin{aligned} 
M 
&=(\frac{4.45}{2.36})^{1.5} \\
M&=2.589 \end{aligned}$$ 
[8]. In either case, the estimate for M is in the order of ~2.6 - ~2.8.  

###Estimation 2 - analyis of crowning thresholds, Sharpsand Creek (ON)
For the second estimation method, we will consider the range of stand types at Sharpsand Creek as treatment to evaluate their effect on crowning thresholds. This site, with many fires in stands that differ primarily by age and treatment, provides the ideal test case for assessing crown fire thresholds in a range of experiments where the primary distinction between stands is in LCBH and ladder fuel density. The value of M can be evaluated using Byram's I_0 with a simple surface ROS model. thresholds applied to the portion of the experimental fire dataset representing the Sharpsand Creek fires.  

```{r Define CFI models for SH stands}
fires.stat <- select(sh.fires, CFI1, CFI2, FFMC, ws, SFC, 
                     MC.SA, ISI, Exp, isi.m, isi.m2) %>%
  ungroup()
#%>%  mutate(CFI2=as.numeric(CFI2))

sh.mod <- glm(CFI1 ~ ISI + Exp, data=fires.stat, family='binomial') #base model, ISI and treatment; both sig. at alpha=0.05

sh.mod2 <- glm(CFI1 ~ isi.m + Exp, data=fires.stat, family='binomial') #ISI using MC.SA, density 2 for TH, 3 for others. Works! p < 0.05 for all

####OK - make sure to change Sharpsand 'partly thinned' plot to MC.SA_dens==3



#Use sh.mod (ISI) for log. regression graph
#Use both for threshold values

#graph with jittered scatter plot
#Isi.seq <- seq(4, 20, by=0.1)

#which model to use? sh.mod with ISI or sh.mod2 with mc.sa or other
use.mod <- sh.mod
#use.mod <- sh.mod2

#generate fake data for log. regression graphs
fdata.IM <- data.frame(fISI=seq(3, 20, by=0.1), fExp='Imm') %>%
  mutate(CFI=predict.glm(use.mod, newdata=list(ISI=fISI, Exp=fExp), 
                         type='response'))
fdata.SM <- data.frame(fISI=seq(3, 20, by=0.1), fExp='Semi-M') %>%
  mutate(CFI=predict.glm(use.mod, newdata=list(ISI=fISI, Exp=fExp), 
                         type='response'))
fdata.TH <- data.frame(fISI=seq(3, 20, by=0.1), fExp='Thin') %>%
  mutate(CFI=predict.glm(use.mod, newdata=list(ISI=fISI, Exp=fExp), 
                         type='response'))

mod.co<-use.mod$coefficients

#solve for p=0.5 in logistical function; 
#p=exp(b0+b1*ISI+b2*ExpSemi-M+b3*ExpThin+c)
#so 0.5=1/(1+exp(b0+b1*ISI+b2*ExpSemi-M+b3*ExpThin+c)
#2=1+exp(b0...)
#1=exp()
#0=b0+b1*ISI+b2*ExpSemi-M+b3*ExpThin+c
#ISI=(b0+b2*ExpSemi-M+b3*ExpThin)/b1

summary(use.mod)  #
```

Based on the logistic regression model, ISI and treatment were both
significant determiners of crown fire success, although the only
significant treatment difference was between immature and thinned
stands.

```{r Log-CFI graph, define functions}
#Main ISI-CFI graph with jittered data

sh.isi.log <- 
  ggplot(fires.stat, aes(y=CFI1, x=isi.m))+
  #scale_shape_manual(values=c(23, 22, 21)) +   #pick nice shapes 
  theme_classic() +
  scale_fill_manual(values=c('red', 'green', 'blue'))+
  scale_shape_manual(values=c(21, 23)) + 
  geom_jitter(aes(fill=Exp, colour=Exp, shape=CFI2), 
              size=2.5, width=0, height = 0.02, alpha=0.4) +  
  geom_line(data=fdata.IM, aes(x=fISI, y=CFI), colour='red') +
  geom_line(data=fdata.SM, aes(x=fISI, y=CFI), colour='green') +
  geom_line(data=fdata.TH, aes(x=fISI, y=CFI), colour='light blue') +
  labs(x='ISI', y='p(CFO)') #+
#  annotate('segment', x=isi.thresh.imm(), xend=isi.thresh.imm(), 
#           y=0.4, yend=0.6, alpha=0.4, colour='red') +
#  annotate('segment', x=isi.thresh.th(), xend=isi.thresh.th(), 
#           y=0.4, yend=0.6, alpha=0.6, colour='light blue') 
  
sh.isi.log

#Solve for I0 with vw77? 
#Turn difference in threshold ISI into I0
#use surface fire spread model
#solve for ROS_0



```

```{r cf thresholds}
#get cf thresholds
isi.thresh.imm <- function(Exp=Imm) { #Imm (NA), Semi-M, or Thin
  isi=-(mod.co[[1]])/mod.co[[2]]
  return(isi)
}

isi.thresh.th <- function(Exp=Thin) { #Imm (NA), Semi-M, or Thin
  isi=-(mod.co[[1]] + mod.co[[4]])/mod.co[[2]]
  return(isi)
}

isi.thresh.semiM <- function(Exp=Semi.M) { #Imm (NA), Semi-M, or Thin
  isi=-(mod.co[[1]] + mod.co[[3]])/mod.co[[2]]
  return(isi)
}


#thresholds with ISI (sh.mod):
#imm: 6.61
#th: 10.31

#thresholds with ISI.mcsa (sh.mod2):
#imm: 4.70
#th: 9.19

#'working mean':
#'imm': 6.3
#'th': 9.8

```


###Validation 3. Comparison of CFO models (with LF scaling, with and without multiplier) with Van Wagner's CFI model, using notional Taylor SF model

##Results
##Discussion
 - Figure showing more complex architecture; 2-story stands, using Eq 2 to estimate contribution of surface fuels to upper stratum crowning
 - Validation 4. Upper canopy crowning
 -Scale issues are obvious here. Following the logic and intent of Van Wagner's (1977) model, the scaling equation represents relationships between canopy strata logically in 1-dimensional space, assuming ladder fuels are positioned beneath overstory canopy fuels. They also may hold at the single tree level, where individual ladder fuel elements (saplings, bark flakes, arboreal lichen) are horizontally juxtaposed with overstory fuels. More detailed survey methods (e.g. Alexander et al. 2004; Reinhardt et al. 2006; Arkin et al. 2023) can describe the frequency distributions associated with vertical strata in a stand, while ecologists have described the spatial patterns of overstory trees in stands across various stand types (Kenkel et al. 1997; Larson and Churchill 2012). Future investigations will be needed on how these spatial distribution patterns affect ladder fuel influences near crown fire initiation thresholds.  

-For the theoretical relationship to hold for actual ladder fuel elements (flammable shrubs, saplings, bark flakes, hanging lichens, etc.), it is imperative that they be connected to surface fuels, and in actual forest settings 

-The concept of a multiplier for comparing elevated fuels to surface fuels is a similar topic to Van Wagner's [-vanwagner1977] 'c' parameter of 0.01, deduced from a comparison of only two experimental fires, as discussed by Alexander [@alexander1998]. Densely-packed litter fuelbeds have lower spread rate than lower density fuelbeds with the same fuel load [@campbell-lochrie2021]. 